A comparison on R tools for nonlinear least squares modeling

### 2023-03-03 RefactoringNLS/ComparingNLStools.Rmd
new submission
update for dimensions of comparison +1

The tools we will compare are the base-R `nls()` function 
and a number of packages that are available on the CRAN repository.

Important:

nlsr
minpack.lm
gslnls
nls2

Worth commentary:

crsnls
NISTnls
nlshelper
nlsic
nlsMicrobio
nlsmsn
nls.multstart
nlstac
nlstools
onls

Mention only:

easynls

@@

@@

Hans Werner Borchers has been helpful in developing the proposal for this project
and in comments on this and related work. Heather Turner co-mentored the project 
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of `selfStart` models and improve package `nlsr` in that respect.
Colin Gillespie (package `benchmarkme`) has been helpful in guiding our attempts
to succinctly summarize computing environments.

>>>>

The tools we will compare are the base-R `nls()` function 
and a number of packages that are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), 
gslnls (@gslnls23), and nls2 (@nls2-22).

Other CRAN packages worth Commentary: ??

onls
crsnls
NISTnls
nlshelper
nlsic
nlsMicrobio
nlsmsn
nls.multstart
nlstac
nlstools
easynls
nlraa
optimx ??

It is likely that there are a number of other tools that could be relevant to
this comparison in the Bioconductor (??ref) collection and on repositories like
Github (??ref) and Gitlab (??ref), but we will not pursue those possibilities 
here.

@@

The tools for nonlinear estimation we consider have been built with different
goals and contexts. We will make our comparison in a number of directions.
We will make a number of specific notes, but our main commentary will consider
the following categories.

### Modeling versus minimizing a sum of squares

The goal of many statisticians and other workers is to estimate a model. The
minimization of a sum of squared residuals -- the nonlinear least squares problem --
is the mechanism to estimate the model. Some of the tools we review focus on the
modeling, so attempt to provide measures of uncertainty and other information. 
The interface to modeling tools will generally try to streamline that task. By
contrast, nonlinear least squares tools are trying to solve the more limited
numerical task.

### Programming language(s)

R tools are our interest, but the tools we consider often use other programming
languages. There are advantages in using a well-established program that has
been much used and tested. Early versions of R were relatively sluggish in 
computational speed. The downside is that we must interface between two or more
programming languages. Changes to one or the other language or its library support
force maintenance issues. Moreover, the pool of expertise with languages like C
and Fortran is diminishing.

### Internal solver stabilization

Early nonlinear least squares programs generally used the Gauss-Newton method 
(?? ref). This is often very effective and efficient, but for a quite large 
proportion of problems can fail to find the solution. Stabilization of the 
computations is relatively straightforward in principle, and overcomes most
of the failures, though sometimes at the cost of more calculations.

The necessarily iterative solution of a nonlinear least squares problem 
to carry out nonlinear modeling requires that we can find smaller values
of a sum of squared residuals. This implies there is a "downhill" direction.
The gradient of the residual sum of squares is computed as the product
of the Jacobian matrix: the partial derivatives of the residual vector 
with respect to the modeling parameters.
(?? do we need to present this, or give a ref)

Most researchers find the effort to interface the differential calculus 
required to specify the Jacobian to a nonlinear estimation tool very 
challenging. The work-around is generally to use a numerical approximation,
but there is the possibility of using symbolic or automatic differentiation,
and there are particular models where workers have provided code for the
Jacobian.

### Output

?? DO WE WANT A COMPARISON TABLE

@@

Hans Werner Borchers has been helpful in developing the proposal for this project
and in comments on this and related work. Heather Turner co-mentored the project 
and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
clarify aspects of `selfStart` models and instigated the "Introduction to nlsr" 
vignette. Colin Gillespie (package `benchmarkme`) has been helpful in guiding our 
attempts to succinctly summarize computing environments.
